Select-based random access \\ to variable-byte encodings

### 2020-06-06 gradu/msc_english_contents.tex
Clarifications

A fundamental method for data compression is variable-length coding \citep{Sal99}. The main idea of variable-length encoding is that 
frequent sequences of data are represented with shorter codewords. Because the sequences of data have different lengths when compressed, it is 
not trivial to determine the exact location of a certain element within the compressed data. If this is required, the usual 
data compression algorithms are inefficient, because data must be decompressed from the beginning to the point where the the sought element is located. 
Fortunately, such random access to elements is not a requirement compression algorithms usually need to fulfill.

@@

\chapter{Rank and Select}
$Rank$ and $select$ are two array operations that are widely used in compressed data structures. $Rank_x$($i$) counts the number of occurences of $x$ before index $i$ in the data. $Select_x$($i$) returns the index of the i-th 
occurence of $x$ in the data. In this thesis, these two operations are assumed to work with a bit array. $Rank_1$($i$) counts the number of bits set to 1 before $i$ and $select_1$($i$) returns the index 
of the i-th bit set to 1. Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. The two operations are related to each other: when used on a bit array 
$B$ = \texttt{0100 1101 \underline{1}011}, $select$(5) returns 8 (underlined) and therefore $rank$(8) returns 4.

@@

VB encoding has several advantages with $select$ and $rank$: the data is compressed in 
blocks of equal length which significantly shrinks the length of $B$. Moreover the bit array $C$ formed from the continuation bits already stores the locations of items and works
as the needed indexing array. In this case, $rank_1$(C, i) would give the number of end blocks before the i-th index and $select_1$(C, i) would give the location of the ending block 
of i-th compressed element.

@@

The $rank$ data structures are based on \citep{Vig08}. A $rank_v$ superblock covers 512 bits and is divided to 512/64 = 8 relative blocks. This means a relative block needs to be able to store a number up to 512 in itself, which takes $\log_2 512$ = 9 bits. 
The first relative block always equals 0, because a relative block counts the number of 1 bits from the start of the superblock. Therefore the remaining 7 relative blocks take 63 bits and fit into one 
64-bit word. For every 512 bits, this implementation requires 64 bits for the superblock and another 64 blocks for the relative blocks, resulting in additional 128/512 = 25\% space.

@@

\chapter{DAC with select query}
\label{chap:select_impl}
Using $select_1$ on the continuation bit array to achieve direct access is more intuitive than using $rank_1$. The element locations are already marked with 1's 
in $B$ and a single $select_1$ query gets the desired starting point, while the aforementioned version \citep{Bri09} used one $rank$ query for each chunk beyond 
the first. Minimizing the amount of $select$ and $rank$ queries is important. They run in constant time but their impact is huge, because rest of the VB decoding 
consists of just a few bit operations.

To use $select_1$ with VB, continuation bits need to be separated from chunks to their own bit array and a $select_1$ structure built over it. Because every compressed element has 1
only on its last block, $Select_1$($i$) returns the location of the end byte of $i$-th element. Therefore the start of  $j$-th element in the block array is at 
block $b_s$ = $select_1$($j$-1) + 1. In this thesis, the implementation was simplified by using only block sizes 4 and 8 to prevent block splitting between bytes.

@@

For example, let the integer bit length be 32 and the original data be compressed to 5 blocks of 4 bits. First, the byte location of the first block is needed. $Select_1$($i$) gets the index location
of needed block, but since the block length differs from byte length, the starting byte location $s$ needs to be calculated $s = (select_1(i) \times 4) \div 8$, the latter operand being for integer division.
The first block might not be at the start of the byte $s$. This is why offset $o = (select_1(i) \times 4)$ $mod$ $8$ is needed (in the example case, $o$ is equal to either 0 or 4). Then a 32-bit word $w$ is read from memory from byte location $s$.

@@

The most intuitive way to calculate the block length of the $i$-th element is to subtract $select_1$(i) from $select_1$(i+1). This however causes an additional 
$select_1$ query, which is costly. A much faster option is to calculate the block length from the continuation bit array. This can be done by reading a word from the bit array and bit-shifting the word 
until the bit representing the start of the $i$-th element is on the rightmost end and then counting the trailing zeros (for example, by using the GCC function \texttt{\_\_builtin\_ctz}). The trailing zeros equal to the number of blocks in the $i$-th element that do not have the continuation bit set. 
This number is then incremented by one to also count the ending block which has the continuation bit set to 1.

@@

After locating and decompressing one value from a compressed data structure, decompressing the subsequent elements cam become significantly easier. Calculating the memory location of the next element may not 
be needed if the next compressed value is located next to the previously fetched element. This is especially beneficial to the select-based VByte decoding, where calculating the start 
byte location takes majority of the runtime. In addition to this, the length of the next compressed integer is very fast to calculate because the required data is already in memory and immediately local to the last point of access.

@@

Table~\ref{table:datasets} describes four different types of synthetic datasets that were used for the experiments. Each dataset was generated to three different sizes (5M, 50M, 500M). 
The content of the datasets were chosen in order to see how different implementations work with different sized data and different magnitudes of integers.
\begin{table}
\centering
\caption{Generated datasets
\label{table:datasets}}
\begin{tabular}{l|l} 
Dataset & Explanation \\
 \hline 
all (5M)   & all integers randomly 1-4 bytes long \\
twolarge   & one eighth of integers 4 bytes long, one eighth 2 bytes long, rest 1 byte \\
onelarge    & one eighth of integers 2 bytes long, rest 4 bits or less \\
onlysmall    & all integers 4 bits or less \\

@@

The results show that the proposed $select$ method outperforms $rank$ when the data set contains several integers that are encoded into more than one block. The difference is explained 
with how $rank$ and $select$ use the data structure. $Select$ is always called once, while $rank$ is called once for each block beyond the first. Figure~\ref{graph:data_comparison} shows average times
of one million $rank$ and $select$ calls with different average block lengths. The effect of multiple $rank$ calls is visible, as is the single $select$ call behavior, which leads to a much flatter curve.

@@

The $rank$ operation can be created with a small number of instructions utilizing a modern \textsc{POPCOUNT} machine instruction \citep{Zho13}. The $select$ approach is 
known to be slower, ranging from 50 to 80 machine instructions. Intel's Haswell CPUs offer two new interesting machine instructions, \textsc{PDEP} and \textsc{TZCNT}. \textsc{PDEP} stores bits
from its first operand based on the location of 1 bits in the second operand. \textsc{TZCNT} is a machine instruction to count trailing zeros. These two can be used to create a $select_1$ query on 
a machine word with just 4 instructions \citep{Pan17a}. They have been tested with the $select$ algorithm of the SDSL-library (which is also used in the proposed method of this thesis) 
and show an improvement of 2-15\% \citep{Pan17b}. Additionally, a trailing zero count function is used to calculate the length of the element. 
Some improvement in speed may happen if the implementation is changed to use \textsc{TZCNT} machine instruction instead.

>>>>

A fundamental method for data compression is variable-length coding \citep{Sal99}. The main idea of variable-length encoding is that 
frequent sequences of data are represented with shorter codewords. Because the sequences of data have different lengths when compressed, it is 
not trivial to determine the exact location of a certain element within the compressed data. If this is required, the usual 
data compression algorithms are inefficient, because data must be decompressed from the beginning to the point where the sought element is located. 
Fortunately, such random access to elements is not a requirement compression algorithms usually need to fulfill.

@@

\chapter{Rank and Select}
$Rank$ and $select$ are two array operations that are widely used in compressed data structures. $Rank_x$($i$) counts the number of occurences of $x$ before index $i$ in the data. $Select_x$($i$) returns the index of the i-th 
occurence of $x$ in the data. In this thesis, these two operations are assumed to work with a bit array. $Rank_1$($i$) counts the number of bits set to 1 before $i$ and $select_1$($i$) returns the index 
of the i-th bit set to 1. Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. The two operations are related to each other: when used on a bit array 
$B$ = \texttt{0100 1101 \underline{1}011}, $select_1$(5) returns 8 (underlined) and therefore $rank_1$(8) returns 4.

@@

VB encoding has several advantages with $select$ and $rank$: the data is compressed in 
blocks of equal length which significantly shrinks the length of $B$. Moreover the bit array $C$ formed from the continuation bits already stores the locations of items and works
as the needed indexing array. In this case, $rank_1$(i) from $C$ would give the number of end blocks before the i-th index and $select_1$(i) would give the location of the ending block 
of i-th compressed element.

@@

The $rank$ data structures are based on \citep{Vig08}. A $rank_v$ superblock covers 512 bits and is divided to 512/64 = 8 relative blocks. This means a relative block needs to be able to store a number up to 512 in itself, which takes $\log_2 512$ = 9 bits. 
The first relative block always equals 0, because a relative block counts the number of 1 bits from the start of the superblock. Therefore the remaining 7 relative blocks take 63 bits and fit into one 
64-bit word. For every 512 bits, this implementation requires 64 bits for the superblock and another 64 bits for the relative blocks, resulting in additional 128/512 = 25\% space.

@@

\chapter{DAC with select query}
\label{chap:select_impl}
Using $select$ on the continuation bit array to achieve direct access is more intuitive than using $rank$. The element locations are already marked with 1's 
in $B$ and a single $select_1$ query gets the desired starting point, while the aforementioned version \citep{Bri09} used one $rank$ query for each chunk beyond 
the first. Minimizing the amount of $select$ and $rank$ queries is important. They run in constant time but their impact is huge, because rest of the VB decoding 
consists of just a few bit operations.

To use $select$ with VB, continuation bits need to be separated from chunks to their own bit array and a $select_1$ structure built over it. Because every compressed element has 1
only on its last block, $Select_1$($i$) returns the location of the end byte of $i$-th element. Therefore the start of  $j$-th element in the block array is at 
block $b_s$ = $select_1$($j$-1) + 1. In this thesis, the implementation was simplified by using only block sizes 4 and 8 to prevent block splitting between bytes.

@@

For example, let the integer bit length be 32 and the original data be compressed to 5 blocks of 4 bits. First, the byte location of the first block is needed. The index location $x$ of the desired block
 is calculated $x = select_1$($i$-1) + 1, but since the block length differs from byte length, the starting byte location $s$ needs to be calculated $s = (x \times 4) \div 8$, the latter operand being for integer division.
The first block might not be at the start of the byte $s$. This is why offset $o = (x \times 4)$ $mod$ $8$ is needed (in the example case, $o$ is equal to either 0 or 4). Then a 32-bit word $w$ is read from memory from byte location $s$.

@@

The most intuitive way to calculate the block length of the $i$-th element is to subtract $select_1$($i$-1) from $select_1$($i$). This however causes an additional 
$select_1$ query, which is costly. A much faster option is to calculate the block length from the continuation bit array. This can be done by reading a word from the bit array and bit-shifting the word 
until the bit representing the start of the $i$-th element is on the rightmost end and then counting the trailing zeros (for example, by using the GCC function \texttt{\_\_builtin\_ctz}). The trailing zeros equal to the number of blocks in the $i$-th element that do not have the continuation bit set. 
This number is then incremented by one to also count the ending block which has the continuation bit set to 1.

@@

After locating and decompressing one value from a compressed data structure, decompressing the subsequent elements can become significantly easier. Calculating the memory location of the next element may not 
be needed if the next compressed value is located next to the previously fetched element. This is especially beneficial to the select-based VByte decoding, where calculating the start 
byte location takes majority of the runtime. In addition to this, the length of the next compressed integer is very fast to calculate because the required data is already in memory and immediately local to the last point of access.

@@

Table~\ref{table:datasets} describes four different types of synthetic datasets that were used for the experiments. Each dataset was generated to three different sizes (5M, 50M, 500M). 
The content of the datasets were chosen in order to see how different implementations work with different sized data and different magnitudes of integers.
\begin{table}
\centering
\caption{Generated datasets
\label{table:datasets}}
\begin{tabular}{l|l} 
Dataset & Explanation \\
 \hline 
all   & all integers randomly 1-4 bytes long \\
twolarge   & one eighth of integers 4 bytes long, one eighth 2 bytes long, rest 1 byte \\
onelarge    & one eighth of integers 2 bytes long, rest 4 bits or less \\
onlysmall    & all integers 4 bits or less \\

@@

The results show that the proposed $select$ method outperforms $rank$ when the data set contains several integers that are encoded into more than one block. The difference is explained 
with how $rank$ and $select$ use the data structure. $Select$ is always called once, while $rank$ is called once for each block beyond the first. Figure~\ref{graph:data_comparison} shows average times
of one million $rank$ and $select$ calls with different average integer lengths. The effect of multiple $rank$ calls is visible, as is the single $select$ call behavior, which leads to a much flatter curve.

@@

The $rank$ operation can be created with a small number of instructions utilizing a modern \textsc{POPCOUNT} machine instruction \citep{Zho13}. The $select$ approach is 
known to be slower, ranging from 50 to 80 machine instructions. Intel's Haswell CPUs offer two new interesting machine instructions, \textsc{PDEP} and \textsc{TZCNT}. \textsc{PDEP} stores bits
from its first operand based on the location of 1 bits in the second operand. \textsc{TZCNT} is a machine instruction to count trailing zeros. These two can be used to create a $select_1$ query on 
a machine word with just 4 instructions \citep{Pan17a}. They have been tested with the $select$ algorithm of the SDSL-library (which is also used in the proposed method of this thesis) 
and show an improvement of 2-15\% \citep{Pan17b}. Additionally, a trailing zero count function is used to calculate the length of the element. 
Some improvement in speed may happen if the implementation is changed to use the \textsc{TZCNT} machine instruction instead.
